---
title: "Segmentation Report"
format:
    html:
        embed-resources: true
        standalone: true
        code-fold: true
        code-tools: true
        fig-width: 12
        fig-height: 8
params:
    sample_name: "sample"
    geojson_file: "inst/extdata/segmentation.geojson"
    whole_cell_only: false
    nuclear_channel: ""
    membrane_channels: ""
    combine_method: ""
    image_file: ""
    run_cellpose: false
    cellpose_diameter: null
    cellpose_min_area: null
    cellpose_flow_threshold: null 
    cellpose_cellprob_threshold: null
    cellpose_model_type: null
    cellpose_pretrained_model: null
    run_mesmer: false
    mesmer_segmentation_level: null
    mesmer_maxima_threshold: null
    mesmer_interior_threshold: null
    mesmer_maxima_smooth: null
    mesmer_min_nuclei_area: null
    mesmer_remove_border_cells: null
    mesmer_pixel_expansion: null
    mesmer_padding: null
    run_cellsam: false
    cellsam_bbox_threshold: null
    cellsam_block_size: null
    cellsam_overlap: null
    cellsam_iou_depth: null
    cellsam_iou_threshold: null
    cellsam_use_wsi: null
    cellsam_gauge_cell_size: null
    cellsam_low_contrast_enhancement: null
    cellsam_model_path: null
    cellsam_min_area: null
---

```{r setup, include=FALSE}
library(devtools)
if (!requireNamespace("spatialVis", quietly = TRUE)) {
    devtools::install_github("WEHI-SODA-Hub/spatialVis")
}
library(spatialVis)
library(dplyr)
```

```{r parse_args, include=FALSE}
parse_value_arg <- function(arg) {
  if (is.null(arg)) {
    return(NA)
  }
  return(arg)
}

sample_name <- params$sample_name
geojson_file <- params$geojson_file
stopifnot(file.exists(geojson_file))
whole_cell_only <- as.logical(params$whole_cell_only)

image_file <- params$image_file

nuclear_channel <- parse_value_arg(params$nuclear_channel)
membrane_channels <- parse_value_arg(params$membrane_channels)

run_cellpose <- as.logical(params$run_cellpose)
cellpose_diameter <- parse_value_arg(params$cellpose_diameter)
cellpose_min_area <- parse_value_arg(params$cellpose_min_area)
cellpose_flow_threshold <- parse_value_arg(params$cellpose_flow_threshold )
cellpose_cellprob_threshold <- parse_value_arg(params$cellpose_cellprob_threshold)
cellpose_model_type <- parse_value_arg(params$cellpose_model_type)
cellpose_pretrained_model <- parse_value_arg(params$cellpose_pretrained_model)

run_mesmer <- as.logical(params$run_mesmer)
mesmer_segmentation_level <- parse_value_arg(params$mesmer_segmentation_level)
mesmer_maxima_threshold <- parse_value_arg(params$mesmer_maxima_threshold)
mesmer_interior_threshold <- parse_value_arg(params$mesmer_interior_threshold)
mesmer_maxima_smooth <- parse_value_arg(params$mesmer_maxima_smooth)
mesmer_min_nuclei_area <- parse_value_arg(params$mesmer_min_nuclei_area)
mesmer_remove_border_cells <- as.logical(params$mesmer_remove_border_cells)
mesmer_pixel_expansion <- parse_value_arg(params$mesmer_pixel_expansion)
mesmer_padding <- parse_value_arg(params$mesmer_padding)

run_cellsam <- as.logical(params$run_cellsam)
cellsam_bbox_threshold <- parse_value_arg(params$cellsam_bbox_threshold)
cellsam_block_size <- parse_value_arg(params$cellsam_block_size)
cellsam_overlap <- parse_value_arg(params$cellsam_overlap)
cellsam_iou_depth <- parse_value_arg(params$cellsam_iou_depth)
cellsam_iou_threshold <- parse_value_arg(params$cellsam_iou_threshold)
cellsam_use_wsi <- as.logical(params$cellsam_use_wsi)
cellsam_gauge_cell_size <- as.logical(params$cellsam_gauge_cell_size)
cellsam_low_contrast_enhancement <- as.logical(params$cellsam_low_contrast_enhancement)
cellsam_model_path <- parse_value_arg(params$cellsam_model_path)
cellsam_min_area <- parse_value_arg(params$cellsam_min_area)
```

```{r detect_measurements, include=FALSE}
# Helper to detect whether compartment/intensity measurement columns are present
.detect_segmentation_measurements <- function(measurement_data,
                                               compartments = c("Cell", "Nucleus"),
                                               compartment_measurements = c("Area", "Length", "Circularity", "Solidity", "Max diameter", "Min diameter"),
                                               intensity_calculations = c("Mean", "Median", "Min", "Max", "Std.Dev")) {
  # Build expected compartment measurement names (e.g. "Cell: Area")
  measurements <- tidyr::expand_grid(
    compartment = compartments,
    measurement = compartment_measurements
  ) %>%
    dplyr::mutate(measurement_colname = paste0(compartment, ": ", measurement)) %>%
    dplyr::pull(measurement_colname)

  # Escape regex-special characters and create regexes anchored at start
  escape_regex <- function(x) {
    escaped <- x
    for (special_character in c("\\", ".", "|", "(", ")", "[", "]", "{", "}", "^", "$", "*", "+", "?")) {
      escaped <- gsub(special_character, paste0("\\\\", special_character), escaped, fixed = TRUE)
    }
    escaped
  }

  measurement_regexes <- vapply(measurements, function(measurement) {
    paste0("^", escape_regex(measurement), "(?:$|\\s)")
  }, character(1))

  has_compartment_measurements <- any(vapply(measurement_regexes, function(re) {
    any(stringr::str_detect(colnames(measurement_data), stringr::regex(re)))
  }, logical(1)))

  # Intensity columns end with ": <Calculation>" where Calculation is one of intensity_calculations
  intensity_re <- paste0(": (", paste(intensity_calculations, collapse = "|"), ")$")
  has_intensity_measurements <- any(grepl(intensity_re, colnames(measurement_data)))

  list(
    has_compartment = has_compartment_measurements,
    has_intensity = has_intensity_measurements
  )
}
```


## Run parameters

```{r, show_params}
params_table <- data.frame(
  Parameter = c("Sample Name", "GeoJSON File", "Nuclear Channel",
                "Membrane Channels", "Image file"),
  Value = c(
    sample_name,
    geojson_file,
    nuclear_channel,
    membrane_channels,
    image_file
  )
)

if (run_cellpose) {
  params_to_add <- data.frame(
    Parameter = c("Method", "Diameter", "Min Area", "Flow Threshold",
                  "Cellprob Threshold", "Model Type", "Pretrained Model"),
    Value = c(
      "Cellpose",
      cellpose_diameter,
      cellpose_min_area,
      cellpose_flow_threshold, 
      cellpose_cellprob_threshold,
      cellpose_model_type,
      cellpose_pretrained_model
    )
  )
  params_table <- rbind(params_table, params_to_add)
} else if (run_mesmer) {
  params_to_add <- data.frame(
    Parameter = c("Method", "Segmentation Level", "Maxima Threshold",
                  "Interior Threshold", "Maxima Smooth", "Min Nuclei Area",
                  "Remove Border Cells", "Pixel Expansion", "Padding"),
    Value = c(
      "Mesmer",
      mesmer_segmentation_level,
      mesmer_maxima_threshold,
      mesmer_interior_threshold,
      mesmer_maxima_smooth,
      mesmer_min_nuclei_area,
      mesmer_remove_border_cells,
      mesmer_pixel_expansion,
      mesmer_padding
    )
  )
  params_table <- rbind(params_table, params_to_add)
} else if (run_cellsam) {
  params_to_add <- data.frame(
    Parameter = c("Method", "Bounding Box Threshold", "Block Size", "Overlap",
                  "IoU Depth", "IoU Threshold", "Use WSI",
                  "Gauge Cell Size", "Low Contrast Enhancement",
                  "Model Path", "Min Area"),
    Value = c(
      "CellSAM",
      cellsam_bbox_threshold,
      cellsam_block_size,
      cellsam_overlap,
      cellsam_iou_depth,
      cellsam_iou_threshold,
      cellsam_use_wsi,
      cellsam_gauge_cell_size,
      cellsam_low_contrast_enhancement,
      cellsam_model_path,
      cellsam_min_area
    )
  )
  params_table <- rbind(params_table, params_to_add)
}

params_table <- rbind(
  params_table,
  data.frame(Parameter = "Whole-cell only", Value = whole_cell_only)
)

knitr::kable(params_table, format = "markdown", col.names = c("Parameter", "Value"))
```

## Segmentation summary statistics

Number of detected cells.

```{r, segmentation_counts}
measurement_data <- get_segmentation_measurements(geojson_file)
if (nrow(measurement_data) == 0) {
  stop("No cells found in GeoJSON file.")
}

object_count <- table(measurement_data$objectType) |> data.frame()
colnames(object_count) <- c("Object", "Count")
knitr::kable(object_count, format = "markdown")
```

Mean and median of measurements across all cells.

```{r, segmentation_stats}
compartments_to_use <- if (whole_cell_only) c("Cell") else c("Cell", "Nucleus")
res <- .detect_segmentation_measurements(
  measurement_data,
  compartments = compartments_to_use
)
if (res$has_compartment) {
  tryCatch({
    summary_table <- summarise_segmentation_stats(
      measurement_data,
      compartments = compartments_to_use,
      other_measurements = if (whole_cell_only) character(0) else c("Nucleus/Cell area ratio")
    )
    knitr::kable(summary_table, format = "markdown")
  }, error = function(e) {
    cat("No measurements found in GeoJSON file.")
  })
} else {
  cat("No measurements found in GeoJSON file.")
}
```

## Plot compartment measurements

Box plots showing measurement distributions across `r if (whole_cell_only) "cellular" else "nuclear and cellular"`
compartments.

```{r, compartment_measurements}
if (res$has_compartment) {
  tryCatch({
    plot_compartment_measurements(
      measurement_data,
      compartments = compartments_to_use
    )
  }, error = function(e) {
    cat("No measurements found in GeoJSON file.")
  })
} else {
  cat("No measurements found in GeoJSON file.")
}
```

## Plot Nucleus/Cell ratio

`r if (whole_cell_only) "Skipped when whole_cell_only is enabled." else "Histogram/density plot of nucleus/cell ratio across cells."`

```{r, plot_nucleus_cell_ratio}
if (whole_cell_only) {
  cat("Skipped nucleus/cell ratio plot because whole_cell_only is enabled.")
} else {
  measurement <- "Nucleus/Cell area ratio"
  col_idx <- which(colnames(measurement_data) == measurement)
  if (length(col_idx) > 0) {
    df <- data.frame(measurement = measurement,
                     value = measurement_data[[col_idx]])

    ggplot2::ggplot(data = df, ggplot2::aes(x = value)) +
      ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)),
                              binwidth = 0.05, colour = "black", fill = "white") +
      ggplot2::geom_density() +
      ggplot2::facet_wrap(~ measurement, scales = "free_y") +
      ggplot2::ggtitle(measurement) +
      ggplot2::theme(axis.title.y = ggplot2::element_blank(),
                     axis.text.x = ggplot2::element_text(angle = 90, hjust = 1),
                     axis.title.x = ggplot2::element_blank(),
                     strip.text.y = ggplot2::element_blank(),
                     panel.background = ggplot2::element_blank())
  } else {
    cat("No nucleus/cell ratio measurements found in GeoJSON file.")
  }
}
```

## Channel intensities

Density plots of median intensity in the cell, cytoplasm, membrane and `r if (whole_cell_only) "cell-associated" else "nuclear"`
compartments.

```{r, channel_intensities}
if (res$has_intensity) {
  tryCatch({
    plot_intensity_measurements(
      measurement_data,
      compartments = if (whole_cell_only) c("Cell", "Membrane") else c("Nucleus", "Cell", "Cytoplasm", "Membrane")
    )
  }, error = function(e) {
    cat("No intensity measurements found in GeoJSON file.")
  })
} else {
  cat("No intensity measurements found in GeoJSON file.")
}
```

## Segmentation preview

Plot of `r if (whole_cell_only) "cell" else "cell and nuclear"` geometries in two random 500x500 pixel regions
containing at least 10 cell objects. Note that any MultiPolygon entities
will be filtered out, so some objects may not be displayed.

```{r, segmentation_preview}
geom_data <- get_segmentation_geometry(geojson_file)
if (whole_cell_only && "nucleus" %in% names(geom_data)) {
  geom_data$nucleus <- NULL
}
min_objects <- 10
image <- NULL
size <- 500

if (file.exists(image_file)) {
  image <- image_file
}

# Use tryCatch here as random selection of area may fail
tryCatch({
  p1 <- plot_geometries(
    geom_data,
    min_objects = min_objects,
    image = image,
    area_width = size,
    area_height = size
  ) |> suppressWarnings()
  p2 <- plot_geometries(
    geom_data,
    min_objects = min_objects,
    image = image,
    area_width = size,
    area_height = size
  ) |> suppressWarnings()
  gridExtra::grid.arrange(p1, p2, ncol = 2)
}, error = function(e) {
  cat("An error occurred:", conditionMessage(e), "\n")
})
```
