---
title: "Segmentation Report"
format:
    html:
        embed-resources: true
        standalone: true
        code-fold: true
        code-tools: true
        fig-width: 12
        fig-height: 8
params:
    sample_name: "sample"
    geojson_file: "inst/extdata/segmentation.geojson"
    nuclear_channel: ""
    membrane_channels: ""
    combine_method: ""
    image_file: ""
    run_cellpose: false
    cellpose_diameter: null
    cellpose_min_area: null
    cellpose_flow_threshold: null 
    cellpose_cellprob_threshold: null
    cellpose_model_type: null
    cellpose_pretrained_model: null
    run_mesmer: false
    mesmer_segmentation_level: null
    mesmer_maxima_threshold: null
    mesmer_interior_threshold: null
    mesmer_maxima_smooth: null
    mesmer_min_nuclei_area: null
    mesmer_remove_border_cells: null
    mesmer_pixel_expansion: null
    mesmer_padding: null
---

```{r setup, include=FALSE}
library(devtools)
if (!requireNamespace("spatialVis", quietly = TRUE)) {
    devtools::install_github("WEHI-SODA-Hub/spatialVis")
}
library(spatialVis)
library(dplyr)
```

```{r parse_args, include=FALSE}
parse_value_arg <- function(arg) {
  if (is.null(arg)) {
    return(NA)
  }
  return(arg)
}

sample_name <- params$sample_name
geojson_file <- params$geojson_file
stopifnot(file.exists(geojson_file))

image_file <- params$image_file

nuclear_channel <- parse_value_arg(params$nuclear_channel)
membrane_channels <- parse_value_arg(params$membrane_channels)

run_cellpose <- as.logical(params$run_cellpose)
cellpose_diameter <- parse_value_arg(params$cellpose_diameter)
cellpose_min_area <- parse_value_arg(params$cellpose_min_area)
cellpose_flow_threshold <- parse_value_arg(params$cellpose_flow_threshold )
cellpose_cellprob_threshold <- parse_value_arg(params$cellpose_cellprob_threshold)
cellpose_model_type <- parse_value_arg(params$cellpose_model_type)
cellpose_pretrained_model <- parse_value_arg(params$cellpose_pretrained_model)

run_mesmer <- as.logical(params$run_mesmer)
mesmer_segmentation_level <- parse_value_arg(params$mesmer_segmentation_level)
mesmer_maxima_threshold <- parse_value_arg(params$mesmer_maxima_threshold)
mesmer_interior_threshold <- parse_value_arg(params$mesmer_interior_threshold)
mesmer_maxima_smooth <- parse_value_arg(params$mesmer_maxima_smooth)
mesmer_min_nuclei_area <- parse_value_arg(params$mesmer_min_nuclei_area)
mesmer_remove_border_cells <- as.logical(params$mesmer_remove_border_cells)
mesmer_pixel_expansion <- params$mesmer_pixel_expansion
mesmer_padding <- parse_value_arg(params$mesmer_padding)
```

```{r detect_measurements, include=FALSE}
# Helper to detect whether compartment/intensity measurement columns are present
.detect_segmentation_measurements <- function(measurement_data,
                                               compartments = c("Cell", "Nucleus"),
                                               compartment_measurements = c("Area", "Length", "Circularity", "Solidity", "Max diameter", "Min diameter"),
                                               intensity_calculations = c("Mean", "Median", "Min", "Max", "Std.Dev")) {
  # Build expected compartment measurement names (e.g. "Cell: Area")
  measurements <- tidyr::expand_grid(
    compartment = compartments,
    measurement = compartment_measurements
  ) %>%
    dplyr::mutate(measurement_colname = paste0(compartment, ": ", measurement)) %>%
    dplyr::pull(measurement_colname)

  # Escape regex-special characters and create regexes anchored at start
  escape_regex <- function(x) {
    escaped <- x
    for (special_character in c("\\", ".", "|", "(", ")", "[", "]", "{", "}", "^", "$", "*", "+", "?")) {
      escaped <- gsub(special_character, paste0("\\\\", special_character), escaped, fixed = TRUE)
    }
    escaped
  }

  measurement_regexes <- vapply(measurements, function(measurement) {
    paste0("^", escape_regex(measurement), "(?:$|\\s)")
  }, character(1))

  has_compartment_measurements <- any(vapply(measurement_regexes, function(re) {
    any(stringr::str_detect(colnames(measurement_data), stringr::regex(re)))
  }, logical(1)))

  # Intensity columns end with ": <Calculation>" where Calculation is one of intensity_calculations
  intensity_re <- paste0(": (", paste(intensity_calculations, collapse = "|"), ")$")
  has_intensity_measurements <- any(grepl(intensity_re, colnames(measurement_data)))

  list(
    has_compartment = has_compartment_measurements,
    has_intensity = has_intensity_measurements
  )
}
```


## Run parameters

```{r, show_params}
params_table <- data.frame(
  Parameter = c("Sample Name", "GeoJSON File", "Nuclear Channel",
                "Membrane Channels", "Image file"),
  Value = c(
    sample_name,
    geojson_file,
    nuclear_channel,
    membrane_channels,
    image_file
  )
)

if (run_cellpose) {
  params_to_add <- data.frame(
    Parameter = c("Method", "Diameter", "Min Area", "Flow Threshold",
                  "Cellprob Threshold", "Model Type", "Pretrained Model"),
    Value = c(
      "Cellpose",
      cellpose_diameter,
      cellpose_min_area,
      cellpose_flow_threshold, 
      cellpose_cellprob_threshold,
      cellpose_model_type,
      cellpose_pretrained_model
    )
  )
  params_table <- rbind(params_table, params_to_add)
} else if (run_mesmer) {
  params_to_add <- data.frame(
    Parameter = c("Method", "Segmentation Level", "Maxima Threshold",
                  "Interior Threshold", "Maxima Smooth", "Min Nuclei Area",
                  "Remove Border Cells", "Pixel Expansion", "Padding"),
    Value = c(
      "Mesmer",
      mesmer_segmentation_level,
      mesmer_maxima_threshold,
      mesmer_interior_threshold,
      mesmer_maxima_smooth,
      mesmer_min_nuclei_area,
      mesmer_remove_border_cells,
      mesmer_pixel_expansion,
      mesmer_padding
    )
  )
  params_table <- rbind(params_table, params_to_add)
}

knitr::kable(params_table, format = "markdown", col.names = c("Parameter", "Value"))
```

## Segmentation summary statistics

Number of detected cells.

```{r, segmentation_counts}
measurement_data <- get_segmentation_measurements(geojson_file)
if (nrow(measurement_data) == 0) {
  stop("No cells found in GeoJSON file.")
}

object_count <- table(measurement_data$objectType) |> data.frame()
colnames(object_count) <- c("Object", "Count")
knitr::kable(object_count, format = "markdown")
```

Mean and median of measurements across all cells.

```{r, segmentation_stats}
res <- .detect_segmentation_measurements(measurement_data)
if (res$has_compartment) {
  tryCatch({
    summary_table <- summarise_segmentation_stats(measurement_data)
    knitr::kable(summary_table, format = "markdown")
  }, error = function(e) {
    cat("No measurements found in GeoJSON file.")
  })
} else {
  cat("No measurements found in GeoJSON file.")
}
```

## Plot compartment measurements

Box plots showing measurement distributions across nuclear and cellular
compartments.

```{r, compartment_measurements}
if (res$has_compartment) {
  tryCatch({
    plot_compartment_measurements(measurement_data)
  }, error = function(e) {
    cat("No measurements found in GeoJSON file.")
  })
} else {
  cat("No measurements found in GeoJSON file.")
}
```

## Plot Nucleus/Cell ratio

Histogram/density plot of nucleus/cell ratio across cells.

```{r, plot_nucleus_cell_ratio}
measurement <- "Nucleus/Cell area ratio"
col_idx <- col_idx <- which(colnames(measurement_data) == measurement)
if (length(col_idx) > 0) {
  df <- data.frame(measurement = measurement,
                   value = measurement_data[[col_idx]])

  ggplot2::ggplot(data = df, ggplot2::aes(x = value)) +
    ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)),
                            binwidth = 0.05, colour = "black", fill = "white") +
    ggplot2::geom_density() +
    ggplot2::facet_wrap(~ measurement, scales = "free_y") +
    ggplot2::ggtitle(measurement) +
    ggplot2::theme(axis.title.y = ggplot2::element_blank(),
                   axis.text.x = ggplot2::element_text(angle = 90, hjust = 1),
                   axis.title.x = ggplot2::element_blank(),
                   strip.text.y = ggplot2::element_blank(),
                   panel.background = ggplot2::element_blank())
} else {
  cat("No nucleus/cell ratio measurements found in GeoJSON file.")
}
```

## Channel intensities

Density plots of median intensity in the cell, cytoplasm, membrane and nuclear
compartments.

```{r, channel_intensities}
if (res$has_intensity) {
  tryCatch({
    plot_intensity_measurements(measurement_data)
  }, error = function(e) {
    cat("No intensity measurements found in GeoJSON file.")
  })
} else {
  cat("No intensity measurements found in GeoJSON file.")
}
```

## Segmentation preview

Plot of cell and nuclear geometries in two random 500x500 pixel regions
containing at least 10 cell objects. Note that any MultiPolygon entities
will be filtered out, so some objects may not be displayed.

```{r, segmentation_preview}
geom_data <- get_segmentation_geometry(geojson_file)
min_objects <- 10
image <- NULL
size <- 500

if (file.exists(image_file)) {
  image <- image_file
}

# Use tryCatch here as random selection of area may fail
tryCatch({
  p1 <- plot_geometries(
    geom_data,
    min_objects = min_objects,
    image = image,
    area_width = size,
    area_height = size
  ) |> suppressWarnings()
  p2 <- plot_geometries(
    geom_data,
    min_objects = min_objects,
    image = image,
    area_width = size,
    area_height = size
  ) |> suppressWarnings()
  gridExtra::grid.arrange(p1, p2, ncol = 2)
}, error = function(e) {
  cat("An error occurred:", conditionMessage(e), "\n")
})
```
